---
title: "Bayesian State Space Model in Stan"
format: 
  html:
    html-math-method: mathjax
    echo: true
    eval: true
---



## State Space Model:

State space models are a generalization of time series models, where the observations are driven by an underlying stochastic transition process, but may be observed with noise.   
We could just use a full Gaussian process model to interpolate the missing (partial) vector values and also propagate the model forward, but that tends to be very slow since the covaraince needs factorizing at every iteration.   See earlier versions of this code if you like inefficient code that converges very slowly.


A state space formulation implies that there is some stochastic process for the mean and we observe some stochastic process.   The observation process allows for the scenario where a Bank doesn't give speeches in a given month ($\alpha = 0$) or something else comes up.  The transition process models the evolution of topic importance, whereas the observation process allows those topics to be observed with noise.  Here "with noise" implies that a central bank only gives a sample of speeches in any time period.




#### Observation process:


Let $(y_{t,k})$ denote the topic proportion for month \(t\) and country \(k\). To keep the model fast enough for routine imputation, we represent each country's latent trajectory with a first-order state-space Gaussian process,

$$f_{1,k} \sim \mathcal{N}\left(\mu_k, \alpha_k^2\right),
\qquad
f_{t,k} \mid f_{t-1,k} \sim \mathcal{N}\left(
\mu_k + \phi_k \left(f_{t-1,k} - \mu_k\right),
\alpha_k^2 \left(1 - \phi_k^2\right)
\right),$$

where $(\phi_k = \exp(-1 / \rho_k)),  $(\rho_k)$ is the temporal length scale for country \(k\), and (\alpha_k)$ is the marginal scale. Marginally, this is still a Gaussian process with covariance


$$\mathrm{Cov}\left(f_{t,k}, f_{t',k}\right)
= \alpha_k^2 \phi_k^{|t-t'|}
= \alpha_k^2 \exp\left(-\frac{|t-t'|}{\rho_k}\right).
$$

This ensures a coherent GP prior while avoiding repeated dense covariance factorizations. Conditional on the latent GP $f_{tk}$ values at month (t), the vector of observations across countries is

$$\mathbf{y}_t \mid \mathbf{f}_t \sim \mathcal{N}_K\left(\mathbf{f}_t, \Sigma\right),
\qquad
\Sigma = \mathrm{diag}(\boldsymbol{\sigma}) \, \Omega \, \mathrm{diag}(\boldsymbol{\sigma}),$$

with correlation matrix $(\Omega)$ between the K countries, and $(\boldsymbol{\sigma})$ country-specific scales. This cross-country covariance enables missing values to be sampled as additional parameters conditional on countries observed in the same month.

If a given month is partitioned into observed countries \(o\) and missing countries \(m\), then the model implies the Gaussian conditional distribution

$$\mathbf{y}_{t,m} \mid \mathbf{y}_{t,o}, \mathbf{f}_t, \Sigma
\sim
\mathcal{N}\left(
\mathbf{f}_{t,m} +
\Sigma_{mo}\Sigma_{oo}^{-1}\left(\mathbf{y}_{t,o} - \mathbf{f}_{t,o}\right),
\;
\Sigma_{mm} - \Sigma_{mo}\Sigma_{oo}^{-1}\Sigma_{om}
\right).$$
This is the usual GP formulation where the mean for the missing values is updated conditional on the deviation from the mean $\mathbf{y}_{t,o} - \mathbf{f}_{t,o}$.  The Covariance for the unobserved locations, $\Sigma_{mm}$, is reduced based on the covariance between missing and observed $\Sigma_{mo}$ scaled by the variance of the observed, $\Sigma_{oo}$. Stan samples the missing entries jointly with the hyperparameters, so posterior draws of the missing values are coherent with the fitted latent process and the learned cross-country covariance structure.

This update differs from a hard-truncated sparse band diagonal kernel. In Stan,  zeroing distant covariances  still leaves a dense-enough covariance matrix that it must be factorized, limiting runtime improvements. The state-space form gives the main speed gain because it scales linearly in the number of time points instead of cubically with the full covariance.

## Basics

Note, when running as a script, we can define `Niter`,  and the saving `filename` externally. Defaults, if not provided, are `Niter = 60000`,  and `filename = "gp_panel_results_inflat.Rdata"`.

```{r version}
#| code-fold: true
#| message: false
#| eval: true
version
library(tidyverse)
library(rlang)
library(lubridate)
library(rstan)
library(coda)
library(reshape2)
library(bayesplot)
options(mc.cores = 4)
rstan_options(auto_write = TRUE)
source("MCMC_functions.R")
```

## User Defined Options

Basic things like the location of data file (**datafolder** ) and its filename  (**datafile**).  The choice of topics (**topic_of_interest**) is another option and some suggested values are provided.   More can be obtained by running:

```{r code suggestion}
#| eval: true
read_csv(datafile) |> pull(anchor)|> unique()
```

Countries to select can also be selected (**countries2use**) and again some options are provided.  Running Canada and the US is fastest.   The G7 countries takes around 45 minutes for 1,000 iterations.   


Finally some other useful options include a file prefix for saving the output (**prefix**), and logical for starting the MCMC from a good place (**use_warm_start**), and either starting points from a previous run if given in (**warm_start_file**), or if left null, reasonable data driven values are selected by infilling missing values through linear interpolation and parameters from basic slice and dice of the data.  Finally the number of iterations is selected.  Using a warm start we can get away with like 5k-10k iterations, which for G7 takes ~4-8 hours to run.
```{r user_defined_options}
#| eval: true
datafolder <- "./"
datafile <- paste0(datafolder, "nmf-time-series-g20.csv")

candidate_topics  <- c("inflat" ,"brexit", "covid", "cbdc", "ukrain" )
topic_of_interest <- candidate_topics[1]

G7_countries <- c(
  "Canada",
  "France",
  "Germany",
  "Italy",
  "Japan",
  "United Kingdom",
  "United States"
)
CanUS <- c(
  "Canada",
  "United States"
)

countries2use <- G7_countries

# start of dataset
start_date = ymd("2008-11-01")
# otherwise the month before the first mention in the dataset.
if(topic_of_interest=="brexit"){
  start_date = ymd("2016-02-01")
}
if(topic_of_interest=="covid"){
  start_date = ymd("2020-01-01")
}
if(topic_of_interest=="cbdc"){
  start_date = ymd("2016-02-01")
}
if(topic_of_interest=="ukrain"){
  start_date = ymd("2014-03-01")
}

prefix <- "codex_gp_"
use_warm_start <- TRUE
warm_start_file <- NULL
Niter <- 5000
```



## Data Loading


From here it's all automated.   Data loading is handled using a helper function.   May as well make a plot of the data while we're here.
```{r dataload}
#| eval: true
data_full <- dataload(datafile, topic_of_interest, countries2use, start_date)
dates_in_use <- data_full$dates_in_use
data <- data_full$data
country_names <- colnames(data)

matplot(
  x = dates_in_use,
  y = data,
  type = "l",
  lty = 1,
  main = topic_of_interest,
  ylab = "Topic Proportion",
  xlab = "Date"
)
legend(
  "topright",
  legend = country_names,
  col = seq_along(countries2use),
  lty = 1,
  cex = 0.7,
  bty = "n"
)
```

## Stan Model

Here's the guts of the model, coded in stan.   All is written out here and then saved as a **.stan** file.  

```{r stan_model}
#| code-fold: true
stan_file <- "stan_gp_panel_missing.stan"

stan_code <- '
data {
  int<lower=1> T;
  int<lower=1> K;
  int<lower=0> N_obs;
  array[N_obs] int<lower=1, upper=T> obs_t;
  array[N_obs] int<lower=1, upper=K> obs_k;
  vector[N_obs] y_obs;
  int<lower=0> N_mis;
  array[N_mis] int<lower=1, upper=T> mis_t;
  array[N_mis] int<lower=1, upper=K> mis_k;
}

parameters {
  vector[K] mu;
  vector<lower=0>[K] alpha;
  vector<lower=0>[K] rho;
  matrix[T, K] z;
  vector<lower=0>[K] sigma_country;
  cholesky_factor_corr[K] L_Omega;
  vector<lower=0>[N_mis] y_mis;
}

transformed parameters {
  matrix[T, K] f;
  matrix[T, K] y_complete;
  matrix[K, K] L_Sigma;
  vector<lower=0, upper=1>[K] phi;

  L_Sigma = diag_pre_multiply(sigma_country, L_Omega);
  y_complete = rep_matrix(0.0, T, K);
  for (k in 1:K) {
    phi[k] = exp(-1 / rho[k]);
  }

  for (n in 1:N_obs) {
    y_complete[obs_t[n], obs_k[n]] = y_obs[n];
  }
  for (n in 1:N_mis) {
    y_complete[mis_t[n], mis_k[n]] = y_mis[n];
  }

  for (k in 1:K) {
    f[1, k] = mu[k] + alpha[k] * z[1, k];
    for (t in 2:T) {
      f[t, k] = mu[k]
        + phi[k] * (f[t - 1, k] - mu[k])
        + alpha[k] * sqrt(1 - square(phi[k])) * z[t, k];
    }
  }
}

model {
  mu ~ normal(0.02, 0.05);
  alpha ~ normal(0, 0.05);
  rho ~ lognormal(log(18), 0.75);
  sigma_country ~ normal(0, 0.05);
  L_Omega ~ lkj_corr_cholesky(2);
  to_vector(z) ~ std_normal();

  for (t in 1:T) {
    vector[K] y_t;
    vector[K] f_t;
    for (k in 1:K) {
      y_t[k] = y_complete[t, k];
      f_t[k] = f[t, k];
    }
    y_t ~ multi_normal_cholesky(f_t, L_Sigma);
  }
}

generated quantities {
  corr_matrix[K] Omega;
  cov_matrix[K] Sigma_country;

  Omega = multiply_lower_tri_self_transpose(L_Omega);
  Sigma_country = quad_form_diag(Omega, sigma_country);
}
'

writeLines(stan_code, stan_file)
cat("Wrote Stan program to", stan_file, "\n")
```

## Prepare For Stan

This is mainly helper functions to deal with finding a strong starting point and passing data to stan as a list.


```{r preamble}
#| code-fold: true
#| eval: true
# if (!exists("Niter")) {
#   Niter <- 60000
# }
# if (!exists("seed")) {
#   seed <- 1234567890
# }
if (!exists("chains")) {
  chains <- 2
}
if (!exists("filename")) {
  filename <- paste0(
    prefix,
    "codex_GP_",
    topic_of_interest,
    "_results_",
    paste0(countries2use, collapse = "_"),
    "_",
    Niter,
    ".Rdata"
  )
}

if (isTRUE(use_warm_start) && is.null(warm_start_file) && file.exists(filename)) {
  warm_start_file <- filename
}

data_for_model <- data
missing_mask <- is.na(data_for_model) | data_for_model == 0
observed_mask <- !missing_mask

obs_idx <- which(observed_mask, arr.ind = TRUE)
mis_idx <- which(missing_mask, arr.ind = TRUE)

stan_data <- list(
  T = nrow(data_for_model),
  K = ncol(data_for_model),
  N_obs = nrow(obs_idx),
  obs_t = if (nrow(obs_idx) > 0) obs_idx[, 1] else integer(0),
  obs_k = if (nrow(obs_idx) > 0) obs_idx[, 2] else integer(0),
  y_obs = if (nrow(obs_idx) > 0) data_for_model[observed_mask] else numeric(0),
  N_mis = nrow(mis_idx),
  mis_t = if (nrow(mis_idx) > 0) mis_idx[, 1] else integer(0),
  mis_k = if (nrow(mis_idx) > 0) mis_idx[, 2] else integer(0)
)

cat("Observed values:", stan_data$N_obs, "\n")
cat("Missing values to impute:", stan_data$N_mis, "\n")
cat("Results file:", filename, "\n")
cat(
  "Warm start source:",
  if (isTRUE(use_warm_start)) {
    if (is.null(warm_start_file)) "data-driven initialization" else warm_start_file
  } else {
    "disabled"
  },
  "\n"
)

fill_missing_panel <- function(x) {
  x_filled <- x
  fallback_value <- median(x[!is.na(x) & x > 0], na.rm = TRUE)
  if (!is.finite(fallback_value)) {
    fallback_value <- 0.02
  }

  for (k in seq_len(ncol(x_filled))) {
    observed <- which(!is.na(x_filled[, k]) & x_filled[, k] > 0)

    if (length(observed) == 0) {
      x_filled[, k] <- fallback_value
    } else if (length(observed) == 1) {
      x_filled[, k] <- x_filled[observed, k]
    } else {
      x_filled[, k] <- approx(
        x = observed,
        y = x_filled[observed, k],
        xout = seq_len(nrow(x_filled)),
        method = "linear",
        rule = 2
      )$y
    }
  }

  x_filled
}

build_data_warm_init <- function(data_for_model, missing_mask, stan_data) {
  filled <- fill_missing_panel(data_for_model)
  T <- nrow(filled)
  K <- ncol(filled)
  mu_init <- colMeans(filled)
  alpha_init <- pmax(apply(filled, 2, sd), 0.01)
  phi_init <- rep(0.95, K)

  for (k in seq_len(K)) {
    centered <- filled[, k] - mu_init[k]
    lagged <- centered[-T]
    current <- centered[-1]
    denom <- sum(lagged^2)

    if (is.finite(denom) && denom > 1e-10) {
      phi_raw <- sum(lagged * current) / denom
      phi_init[k] <- min(max(phi_raw, 0.80), 0.995)
    }
  }

  rho_init <- pmax(-1 / log(phi_init), 0.5)
  innovation_scale <- alpha_init * sqrt(pmax(1 - phi_init^2, 1e-6))
  z_init <- matrix(0, nrow = T, ncol = K)

  for (k in seq_len(K)) {
    z_init[1, k] <- (filled[1, k] - mu_init[k]) / alpha_init[k]
    for (t in 2:T) {
      z_init[t, k] <- (
        filled[t, k] -
          mu_init[k] -
          phi_init[k] * (filled[t - 1, k] - mu_init[k])
      ) / innovation_scale[k]
    }
  }

  y_mis_init <- if (stan_data$N_mis > 0) {
    pmax(filled[missing_mask], 1e-6)
  } else {
    numeric(0)
  }

  list(
    mu = mu_init,
    alpha = alpha_init,
    rho = rho_init,
    z = z_init,
    sigma_country = pmax(alpha_init * 0.25, 0.01),
    L_Omega = diag(K),
    y_mis = y_mis_init
  )
}

load_saved_fit <- function(path) {
  file_ext <- tolower(tools::file_ext(path))

  if (file_ext == "rds") {
    fit_obj <- readRDS(path)
  } else {
    fit_env <- new.env(parent = emptyenv())
    loaded_names <- load(path, envir = fit_env)
    stanfit_names <- loaded_names[
      vapply(
        loaded_names,
        function(nm) inherits(fit_env[[nm]], "stanfit"),
        logical(1)
      )
    ]

    if (length(stanfit_names) == 0) {
      stop("No stanfit object found in ", path)
    }
    fit_obj <- fit_env[[stanfit_names[1]]]
  }

  if (!inherits(fit_obj, "stanfit")) {
    stop(path, " does not contain a stanfit object.")
  }

  fit_obj
}

build_fit_warm_init <- function(path, stan_data, data_for_model, missing_mask) {
  fit_obj <- load_saved_fit(path)
  draws <- rstan::extract(
    fit_obj,
    pars = c("mu", "alpha", "rho", "z", "sigma_country", "L_Omega", "y_mis")
  )
  required_pars <- c("mu", "alpha", "rho", "z", "sigma_country", "L_Omega")

  missing_required <- setdiff(required_pars, names(draws))
  if (length(missing_required) > 0) {
    stop(
      "Warm start fit is missing required parameters: ",
      paste(missing_required, collapse = ", ")
    )
  }

  n_draws <- dim(draws$mu)[1]
  draw_id <- n_draws
  fallback_init <- build_data_warm_init(data_for_model, missing_mask, stan_data)

  if (
    length(draws$mu[draw_id, ]) != stan_data$K ||
      ncol(draws$z[draw_id, , ]) != stan_data$K ||
      nrow(draws$z[draw_id, , ]) != stan_data$T
  ) {
    stop("Warm start fit dimensions do not match the current panel.")
  }

  y_mis_init <- fallback_init$y_mis
  if (stan_data$N_mis > 0 && "y_mis" %in% names(draws)) {
    if (length(draws$y_mis[draw_id, ]) == stan_data$N_mis) {
      y_mis_init <- pmax(draws$y_mis[draw_id, ], 1e-6)
    }
  }

  list(
    mu = as.numeric(draws$mu[draw_id, ]),
    alpha = pmax(as.numeric(draws$alpha[draw_id, ]), 1e-4),
    rho = pmax(as.numeric(draws$rho[draw_id, ]), 0.1),
    z = draws$z[draw_id, , ],
    sigma_country = pmax(as.numeric(draws$sigma_country[draw_id, ]), 1e-4),
    L_Omega = draws$L_Omega[draw_id, , ],
    y_mis = y_mis_init
  )
}

jitter_warm_init <- function(init_list, chain_id) {
  jitter_scale <- 0.02 * chain_id

  list(
    mu = init_list$mu + rnorm(length(init_list$mu), 0, 0.01 + jitter_scale / 2),
    alpha = pmax(
      init_list$alpha * exp(rnorm(length(init_list$alpha), 0, jitter_scale)),
      1e-4
    ),
    rho = pmax(
      init_list$rho * exp(rnorm(length(init_list$rho), 0, jitter_scale)),
      0.1
    ),
    z = init_list$z + matrix(
      rnorm(length(init_list$z), 0, 0.05 + jitter_scale),
      nrow = nrow(init_list$z),
      ncol = ncol(init_list$z)
    ),
    sigma_country = pmax(
      init_list$sigma_country * exp(
        rnorm(length(init_list$sigma_country), 0, jitter_scale)
      ),
      1e-4
    ),
    L_Omega = init_list$L_Omega,
    y_mis = pmax(
      init_list$y_mis * exp(rnorm(length(init_list$y_mis), 0, jitter_scale)),
      1e-6
    )
  )
}

make_init_function <- function(
  use_warm_start,
  warm_start_file,
  data_for_model,
  missing_mask,
  stan_data
) {
  if (!isTRUE(use_warm_start)) {
    return("random")
  }

  base_init <- NULL

  if (!is.null(warm_start_file) && file.exists(warm_start_file)) {
    base_init <- tryCatch(
      build_fit_warm_init(
        path = warm_start_file,
        stan_data = stan_data,
        data_for_model = data_for_model,
        missing_mask = missing_mask
      ),
      error = function(e) {
        message("Warm start from file failed: ", conditionMessage(e))
        NULL
      }
    )
  }

  if (is.null(base_init)) {
    base_init <- build_data_warm_init(data_for_model, missing_mask, stan_data)
    message("Using data-driven warm start.")
  } else {
    message("Using posterior warm start from ", warm_start_file)
  }

  function(chain_id = 1) {
    jitter_warm_init(base_init, chain_id)
  }
}

init_fun <- make_init_function(
  use_warm_start = use_warm_start,
  warm_start_file = warm_start_file,
  data_for_model = data_for_model,
  missing_mask = missing_mask,
  stan_data = stan_data
)
```

## Run MCMC via Stan

Stan uses the NUTS variant of Hamiltonian Monte Carlo. With this formulation, it's quite efficient, yet still slow since each of the **k** countries need their own missing values and covariances.

```{r MCMC}
#| code-fold: true
#| eval: false
T1 <- Sys.time()

fit <- stan(
  file = stan_file,
  data = stan_data,
  init = init_fun,
  iter = Niter,
  warmup = floor(Niter / 2),
  chains = chains,
  # seed = seed, # i don't generally like using a fixed seed, since then it is just a deterministic process.
  algorithm = "NUTS",
  refresh = max(1, floor(Niter / 20)),
  control = list(
    adapt_delta = 0.95,
    max_treedepth = 12
  )
)

elapsed <- Sys.time() - T1
cat(
  paste(
    "total compute time:",
    round(as.numeric(elapsed, units = "hours"), 2),
    "hours.\n"
  )
)
# save.image("5k iters.Rdata")
```

## Posterior Summaries


```{r data_load_for_rendering}
#| code-fold: true
#| eval: true

# load some data if it exists.
# this is useful if you want to render this document.  It'll load the appropriate filename and go from there.

if(filename %in% list.files( pattern = filename)){
  load(filename)
}

```



```{r posterior_summaries}
#| eval: true
#| code-fold: true
fit_summary <- summary(
  fit,
  pars = c("mu", "alpha", "rho", "sigma_country")
)$summary |>
  as_tibble(rownames = "parameter")

print(fit_summary)

if (stan_data$N_mis > 0) {
  mis_pars <- paste0("y_mis[", seq_len(stan_data$N_mis), "]")
  mis_summary <- summary(fit, pars = mis_pars)$summary |>
    as_tibble(rownames = "parameter") |>
    mutate(
      date = dates_in_use[stan_data$mis_t],
      country = country_names[stan_data$mis_k]
    ) |>
    dplyr::select(
      date,
      country,
      mean,
      sd,
      `2.5%`,
      `25%`,
      `50%`,
      `75%`,
      `97.5%`,
      n_eff,
      Rhat
    )

  print(mis_summary, n = min(25, nrow(mis_summary)))

  posterior_mean_matrix <- data_for_model
  posterior_mean_matrix[missing_mask] <- mis_summary$mean

  posterior_plot_df <- posterior_mean_matrix |>
    as.data.frame() |>
    setNames(country_names) |>
    mutate(date = dates_in_use) |>
    pivot_longer(
      cols = all_of(country_names),
      names_to = "country",
      values_to = "posterior_mean"
    )

  imputed_plot_df <- mis_summary |>
    rename(
      lower_95 = `2.5%`,
      median_value = `50%`,
      upper_95 = `97.5%`
    )

  ggplot() +
    geom_linerange(
      data = imputed_plot_df,
      aes(ymin = lower_95, ymax = upper_95, x = date, color = country, group = country),
      linewidth = 0.5,
      alpha = 0.5,
      show.legend = FALSE
    ) +
    geom_point(
      data = imputed_plot_df,
      aes(y = mean, x = date, color = country, group = country),
      shape = 21,
      fill = "white",
      stroke = 0.9,
      size = 2.2,
      show.legend = FALSE
    ) +
    geom_line(
      data = posterior_plot_df,
      aes(x = date, y = posterior_mean, color = country, group = country),
      linewidth = 0.8, alpha = 0.9) +
    labs(
      title = paste(topic_of_interest, "with posterior mean imputations"),
      subtitle = "Imputed points are marked with 95% posterior intervals",
      x = "Date",
      y = "Topic Proportion",
      color = "Country"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      legend.position = "right",
      panel.grid.minor = element_blank()
    )
}
```




### Plot some basic disgnostics such as traceplots:


```{r traceplots}
#| eval: true
#| code-fold: true
plot(extract(fit, "lp__")[[1]],
     ylab = "un-normalized log posterior")


matplot(extract(fit, "mu")[[1]], type = 'l', main = "mu")
pairs(extract(fit, "mu")[[1]], main = "mu")

pairs(extract(fit, "rho")[[1]], main = "rho")


pairs(extract(fit, "alpha")[[1]], main = "alpha")

```



### Show data fit

```{r show_data_fit}
#| eval: true
#| code-fold: true
fitted_mean_matrix <- data_for_model
if (exists("mis_summary")) {
  fitted_mean_matrix[missing_mask] <- mis_summary$mean
}

f_draws <- rstan::extract(fit, pars = "f")$f
fitted_lower_matrix <- apply(f_draws, c(2, 3), quantile, probs = 0.025)
fitted_upper_matrix <- apply(f_draws, c(2, 3), quantile, probs = 0.975)

fitted_plot_df <- fitted_mean_matrix |>
  as.data.frame() |>
  setNames(country_names) |>
  mutate(date = dates_in_use) |>
  pivot_longer(
    cols = all_of(country_names),
    names_to = "country",
    values_to = "fitted_mean"
  )

fitted_interval_df <- tibble(
  date = rep(dates_in_use, times = length(country_names)),
  country = rep(country_names, each = length(dates_in_use)),
  fitted_lower_95 = as.vector(fitted_lower_matrix),
  fitted_upper_95 = as.vector(fitted_upper_matrix)
)

observed_plot_df <- data_for_model |>
  as.data.frame() |>
  setNames(country_names) |>
  mutate(date = dates_in_use) |>
  pivot_longer(
    cols = all_of(country_names),
    names_to = "country",
    values_to = "observed_value"
  ) |>
  filter(!is.na(observed_value), observed_value != 0) |>
  mutate(
    t_index = match(date, dates_in_use),
    k_index = match(country, country_names)
  ) |>
  rowwise() |>
  mutate(
    fitted_lower_95 = quantile(f_draws[, t_index, k_index], 0.025),
    fitted_median = median(f_draws[, t_index, k_index]),
    fitted_upper_95 = quantile(f_draws[, t_index, k_index], 0.975)
  ) |>
  ungroup()

ggplot() +
  geom_ribbon(
    data = fitted_interval_df,
    aes(
      x = date,
      ymin = fitted_lower_95,
      ymax = fitted_upper_95,
      group = country
    ),
    alpha = 0.2,
    fill = "black",
    color = "black",
    show.legend = FALSE
  ) +
  geom_line(
    data = fitted_plot_df,
    aes(x = date, y = fitted_mean, color = country, group = country),
    linewidth = 0.9,
    alpha = 0.9
  ) +
  # geom_linerange(
  #   data = observed_plot_df,
  #   aes(
  #     x = date,
  #     ymin = fitted_lower_95,
  #     ymax = fitted_upper_95,
  #     color = country
  #   ),
  #   linewidth = 0.5,
  #   alpha = 0.45,
  #   show.legend = FALSE
  # ) +
  geom_point(
    data = observed_plot_df,
    aes(x = date, y = observed_value, color = country),
    size = 1.3,
    alpha = 0.75
  ) +
  geom_line(
    data = observed_plot_df,
    aes(x = date, y = fitted_median),
    colour = "black",
    size = 1.8,
    show.legend = FALSE
  ) +
  labs(
    title = paste(topic_of_interest, "model fit to observed data"),
    subtitle = paste(
      "Lines show fitted posterior means;",
      "observed points are overlaid with 95% fitted intervals"
    ),
    x = "Date",
    y = "Topic Proportion",
    color = "Country"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )+
  facet_wrap(~country)
```




## Correlation matrix Omega

```{r omega_plots}
#| eval: true
#| code-fold: true
omega_draws <- rstan::extract(fit, pars = "Omega")$Omega

omega_median <- apply(omega_draws, c(2, 3), median)
omega_variance <- apply(omega_draws, c(2, 3), var)

dimnames(omega_median) <- list(country_names, country_names)
dimnames(omega_variance) <- list(country_names, country_names)

omega_plot_df <- function(mat, value_name) {
  expand.grid(
    country_row = country_names,
    country_col = country_names,
    stringsAsFactors = FALSE
  ) |>
    mutate(
      value = as.vector(mat),
      country_row = factor(country_row, levels = rev(country_names)),
      country_col = factor(country_col, levels = country_names)
    ) |>
    rename(!!value_name := value)
}

plot_omega_heatmap <- function(plot_data, fill_var, title_text, midpoint = NULL) {
  p <- ggplot(
    plot_data,
    aes(x = country_col, y = country_row, fill = .data[[fill_var]])
  ) +
    geom_tile(color = "white", linewidth = 0.4) +
    coord_equal() +
    labs(
      title = title_text,
      x = NULL,
      y = NULL,
      fill = fill_var
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.text.y = element_text(hjust = 1)
    )

  if (!is.null(midpoint)) {
    p <- p + scale_fill_gradient2(
      low = "#2166ac",
      mid = "#f7f7f7",
      high = "#b2182b",
      midpoint = midpoint,
      limits = c(-1, 1)
    )
  } else {
    p <- p + scale_fill_gradient(
      low = "#f7fbff",
      high = "#08306b"
    )
  }

  p
}

omega_median_df <- omega_plot_df(omega_median, "median_correlation")
omega_variance_df <- omega_plot_df(omega_variance, "correlation_variance")

omega_median_plot <- plot_omega_heatmap(
  omega_median_df,
  fill_var = "median_correlation",
  title_text = "Posterior median of Omega",
  midpoint = 0
)

omega_variance_plot <- plot_omega_heatmap(
  omega_variance_df,
  fill_var = "correlation_variance",
  title_text = "Posterior variance of Omega"
)

print(omega_median_plot)
print(omega_variance_plot)

output_stub <- tools::file_path_sans_ext(filename)
omega_median_file <- paste0(output_stub, "_Omega_median.png")
omega_variance_file <- paste0(output_stub, "_Omega_variance.png")

ggsave(
  filename = omega_median_file,
  plot = omega_median_plot,
  width = 8,
  height = 6,
  dpi = 300
)

ggsave(
  filename = omega_variance_file,
  plot = omega_variance_plot,
  width = 8,
  height = 6,
  dpi = 300
)

cat("Saved Omega median image to", omega_median_file, "\n")
cat("Saved Omega variance image to", omega_variance_file, "\n")
```

## Save Output

```{r save_output}
#| code-fold: true

save(
  fit,
  stan_data,
  data_for_model,
  dates_in_use,
  country_names,
  topic_of_interest,
  file = filename
)
cat("Saved fit and data objects to", filename, "\n")
```



## Running this on real data:

- Code could be made more efficient using Sequential Monte Carlo. Stan is generally pretty good though.
- Another reasonable solution is to go fully frequentist and optimize the marginal likelihood by integrating over the nuissance states **X** using Integrated Nested Laplace Approximation.  The package **tmb** is designed to do this quickly.
- Use the script to extract the R code, because this will take a while to run. 

```{r purl}
#| eval: false

knitr::purl("DATA_stan_state_space_model.qmd")


```
